---
title: "Regression"
output: 
  html_document: 
    theme: readable
    css: ../css/custom.css
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(corrplot)
data(mtcars)
knitr::opts_chunk$set(echo = TRUE)
set.seed(10108)
```

---


## Dataset

A data frame with 32 observations on 11 variables.

| var  | Description                              |
|------|------------------------------------------|
| mpg  | Miles/(US) gallon                        |
| cyl  | Number of cylinders                      |
| disp | Displacement (cu.in.)                    |
| hp   | Gross horsepower                         |
| drat | Rear axle ratio                          |
| wt   | Weight (1000 lbs)                        |
| qsec | 1/4 mile time                            |
| vs   | V/S                                      |
| am   | Transmission (0 = automatic, 1 = manual) |
| gear | Number of forward gears                  |
| carb | Number of carburetors                    |

```{r mtcars}
head(mtcars)
```

```{r mtcarsplot, fig.width=4, fig.height=4, echo=FALSE}
corrplot(cor(mtcars[, -1]), method="ellipse")
```
<br/>


## OLS Regression

Given a data set ${\{y_{i},\,x_{i1},\ldots ,x_{ip}\}_{i=1}^{n}}$ of n statistical units, the model takes the form:
$$
y_{i}=\beta _{1}x_{i1}+\cdots +\beta _{p}x_{ip}+\varepsilon _{i}=\mathbf {x} _{i}^{\rm {T}}{\boldsymbol {\beta }}+\varepsilon _{i},\qquad i=1,\ldots ,n,
$$
where,  
$y_{i}$ is the response variable,  
$x_{i1},\,x_{i2},\,\ldots ,\,x_{ip}$ are the regressors,  
$\beta$ is a p-dimensional parameter vector (estimated by least-squares method)  
$\epsilon_{i}$ is the error term.  

### Fitting the Model

We will regress the *mpg* variable onto *disp, cyl, am*. A linear model is fitted by least-squares, which minimizes the sum of squared residuals and  leads to a closed-form expression for the estimated value of the unknown parameter $\boldsymbol \beta$.  

Objective:
$$
{\hat {\boldsymbol {\beta }}}={\rm {arg}}\min _{\boldsymbol \beta }\,\left\lVert \mathbf {y}-\mathbf {X}{\boldsymbol \beta } \right\rVert ,
$$

which leads to:
$$
{\hat {\boldsymbol {\beta }}}=(\mathbf {X} ^{\rm {T}}\mathbf {X} )^{-1}\mathbf {X} ^{\rm {T}}\mathbf {y} ={\big (}\,{\textstyle \sum }\mathbf {x} _{i}\mathbf {x} _{i}^{\rm {T}}\,{\big )}^{-1}{\big (}\,{\textstyle \sum }\mathbf {x} _{i}y_{i}\,{\big )}.
$$

```{r createmodel}
model <- lm(mpg ~ cyl + disp + am, data = mtcars)
model$coefficients
```

```{r summary, echo=FALSE}
model.summ <- summary(model)
```

Adjusted R-squared: `r model.summ$adj.r.squared`  
F-statistic: `r model.summ$fstatistic[1]`

```{r residualplot, echo=FALSE, fig.width=5, fig.height=4}
source("lmplots.R")
diagPlot(model)$rvfPlot
```

```{r predictsales}
# sales vs predicted sales
head(data.frame(mpg = mtcars$mpg, pred = predict(model)), 10)
```
<br/>
  
  
## LOESS

Suppose a dataset consists of $N$ pairs of observations, $\{(x_i, y_i)\}_{i=1}^N$. For fitting a point $x$, define a bandwidth $h(x)$ and a smoothing window $(x−h(x),\,x+h(x))$. The LOWESS algorithm finds the vector $\boldsymbol\beta$ of coefficients which minimizes the locally weighted sum of squares
$$
\boldsymbol{\hat\beta} = \underset {\boldsymbol\beta \in \mathbb R^{p + 1}}{\arg\min} \sum_{i=1}^N w_i(x)(y_i - \langle{\boldsymbol\beta, \, A(x_i - x)}\rangle)^2
$$
where,  
$$
\begin{align}
A_0(v) = 1, &&
A_k(v) = \frac {v^k}{k!}, && 
w_i(x) = W\left(\frac {x_i - x}{h(x)}\right),
\end{align}
$$
$W(u)$ is a weight function that assigns largest weight to observations close to $x$.  

The local regression estimate is
$$
\hat f(x) = \langle \boldsymbol {\hat\beta}, \, A(0) \rangle = \boldsymbol {\hat\beta}_0.
$$


```{r loess}
set.seed(19)

period <- 120
x <- 1:period
y <- sin(2*pi*x/period) + runif(length(x),-1,1)

spanlist <- c(0.10, 0.25, 0.50, 0.75, 1.00, 2.00)
loess.lines <- sapply(spanlist, function(s) {
  y.loess <- loess(y ~ x, span = s, data.frame(x = x, y = y))
  cat("span = ", format(s, nsmall = 2), ", rse: ", y.loess$s, "\n", sep = "")
  predict(y.loess)
})
```
<br/>  
  
```{r loessplot, echo=FALSE, fig.width=7, fig.height=4}
colnames(loess.lines) <- paste("span =", spanlist)
loess.plt <- stack(as.data.frame(loess.lines))
loess.plt <- cbind(loess.plt, x = x)  # recycle x
ggplot(loess.plt, aes(x = x, y = values, group = ind, color = ind)) +
  geom_line(size = 0.5) +
  geom_point(aes(x, y), data = data.frame(x, y), inherit.aes = F) +
  scale_color_discrete("") +
  ggtitle("LOESS Smoothing") +
  theme_bw()
```
<br/>


## Logistic Regression

Formally, the logistic regression model is that
$$
\log {\frac {p(x)}{1 - p(x)}} = \beta_0 + \beta^T x.
$$
Suppose that we are given a sample $(x_i, y_i)$, $i = 1,\ldots,n$, where $y_i$ denotes the class $\in \{1, 2\}$ of the $i$-th observation. Then, assuming that the class labels are conditionally independent, the log-likelihood is given by
$$
\begin{aligned}
\ell(\beta_0, \beta) &= \sum_{i=1}^n \log {P(C = y_i \mid X = x_i)} &\\
                     &= \sum_{i=1}^n \left\{u_i \cdot (\beta_0 + \beta^T x_i) - \log {(1 + \exp (\beta_0 + \beta^T x_i))} \right\}
\end{aligned}
$$
where, 
$$
\begin{align}
u_i = \begin{cases} 1 & {\mbox{if }} y_i = 1 \\
                    0 & {\mbox{if }} y_i = 2 \end{cases} 
\end{align}
$$

The coefficients are estimated by maximizing the likelihood,
$$
\hat\beta_0, \hat\beta = \underset {\beta_0 \in \mathbb R, \beta \in \mathbb R^p}{\arg \max} \ell(\beta_0, \beta)
$$

```{r logreg}
mtcars.glm <- glm(vs ~ mpg, data = mtcars, family = binomial)
mtcars.glm
```
<br/>  
  
```{r logregplot, echo=FALSE, fig.width=5, fig.height=3.5}
ggplot(mtcars, aes(x = mpg, y = vs)) +
  theme_bw(base_size=16) +
  geom_point(alpha = 0.5, position = position_jitter(w = 0, h = 0.02)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se = TRUE, colour = 'red', alpha = 0.3, fill = "grey") +
  labs(y = "Probability", x = "mpg")
```



